---
title: "American Lobster: Where Will They Go?"
subtitle: "Species Distribution Modeling Under Climate Change"
author: "Paung Khuat (PK)"
date: "January 27, 2026"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE, 
  message = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5
)
source("/home/pkhuat28/ColbyForecasting/setup.R")
SPECIES = "Homarus americanus"
BEST_MODEL = "default_maxent"

theme_map <- function() {
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    plot.title = element_text(size = 10)
  )
}
```

## My Species: American Lobster

The **American Lobster** (*Homarus americanus*) is very important in the Gulf of Maine.

**Why I chose this species:**

- Most valuable fishery in New England
- Climate change is already affecting them
- Fishermen want to know: where will lobsters go?
- "Maine & Lobster" conversation

## My Question

<div class="centered" style="font-size: 1.5em; margin-top: 100px;">
**How will climate change affect American Lobster habitat?**
</div>

<div class="centered" style="margin-top: 50px;">
I used **Species Distribution Modeling (SDM)** to answer this.
</div>

## Step 1: Get the Data

I downloaded  **209,167** lobster records from OBIS.

After cleaning (removing missing values, filtering by year), I kept **104,630** good quality records (Some had missing dates.Some had missing individual counts. Some were outside the Brickman model area. Some were before 1970
)

```{r load_data}
obs = read_obis(SPECIES) |>
  filter(!is.na(eventDate), !is.na(individualCount)) |>
  mutate(month = factor(month, levels = month.abb)) |>
  filter(year >= 1970)

db = brickman_database() |> filter(scenario == "STATIC", var == "mask")
mask = read_brickman(db)
hitOrMiss = extract_brickman(mask, obs)
obs = obs |> filter(!is.na(hitOrMiss$value))
coast = read_coastline()
```

## Step 2: Where Are the Lobsters?

```{r obs_map, fig.width = 10, fig.height = 5.5}
ggplot() +
  geom_sf(data = obs, alpha = 0.3, size = 0.3, color = "darkblue") +
  geom_sf(data = coast, col = "orange", linewidth = 0.3) +
  facet_wrap(~month, ncol = 6) +
  theme_map() +
  theme(strip.text = element_text(size = 8))
```

**What I see:** Lobsters are mostly in **coastal areas** near Massachusetts and Maine.

## Step 3: What Environment Do Lobsters Like?

```{r env_prep}
model_input = read_model_input(scientificname = SPECIES, version = "v1")
present = read_brickman(add = c("depth"))
variables = extract_brickman(present, model_input, form = "wide") |>
  mutate(class = model_input$class) |>
  select(-.id)
```

```{r env_plot, fig.height = 4.5}
plot_pres_vs_bg(variables |> select(-month), "class")
```

## What Lobsters Prefer

Based on the data:

| Factor | What Lobsters Like |
|--------|-------------------|
| **Depth** | Shallow water (< 250m) |
| **Bottom Temp** | Cool (5-10°C) |
| **Surface Temp** | Not too warm |

Lobsters avoid deep water and very warm temperatures!

## Step 4: Train the Model

I trained 4 models and compared their confusion matrices:

| Model | Accuracy | Found Lobsters? | Problem |
|-------|----------|-----------------|---------|
| GLM | 85.4% | 0 out of 1244 | Predicts all "no lobster" |
| Boosted Trees | 85.4% | 0 out of 1244 | Predicts all "no lobster" |
| Random Forest | 84.9% | 2 out of 1244 | Almost all "no lobster" |
| **MaxEnt** | 57.2% | **1033 out of 1244** | **Actually finds lobsters!** |

## Why I Chose MaxEnt

- Other models have higher accuracy BUT they cheat!
- They just say "no lobster everywhere" (easy to be right most of the time)
- **MaxEnt correctly identifies 83% of lobster locations**
- For species distribution, finding presence is more important!


## Step 5: Present Day Prediction

```{r nowcast, fig.width = 10, fig.height = 5.5, results='hide'}
cfg = read_configuration(scientificname = SPECIES, version = "v1", path = data_path("models"))
db = brickman_database()

present_conditions = read_brickman(db |> filter(scenario == "PRESENT", interval == "mon"),
                                   add = c("depth", "month")) |>
  select(all_of(cfg$keep))

model_fits = read_model_fit(filename = "Homarus_americanus-v1-model_fits")
nowcast = predict_stars(model_fits, present_conditions)

plot_prediction(nowcast[BEST_MODEL]) + 
  ggtitle("Present Day: Where Lobsters Can Live") +
  theme_map() +
  theme(strip.text = element_text(size = 8))
```

## What the Present Map Shows

- **Winter/Spring (Jan-Apr):** Good habitat across the shelf (orange)
- **Summer (May-Aug):** Lobsters leave warm areas (purple)
- **Fall (Sep-Dec):** Lobsters come back

This matches real lobster behavior!

## Step 6: Future Prediction (RCP 8.5, Year 2075)

```{r rcp85_2075, fig.width = 10, fig.height = 5.5, results='hide'}
covars_rcp85_2075 = read_brickman(db |> filter(scenario == "RCP85", year == 2075, interval == "mon"),
                                  add = c("depth", "month")) |>
  select(all_of(cfg$keep))
forecast_rcp85_2075 = predict_stars(model_fits, covars_rcp85_2075)

plot_prediction(forecast_rcp85_2075[BEST_MODEL]) + 
  ggtitle("Year 2075 (High Emissions): Future Lobster Habitat") +
  theme_map() +
  theme(strip.text = element_text(size = 8))
```

## Compare: Now vs 2075

```{r compare, fig.width = 10, fig.height = 5,results='hide'}
library(patchwork)

p1 = plot_prediction(nowcast[BEST_MODEL]) + 
  ggtitle("PRESENT") + 
  theme_map() +
  theme(strip.text = element_text(size = 7), legend.position = "none")

p2 = plot_prediction(forecast_rcp85_2075[BEST_MODEL]) + 
  ggtitle("2075 (RCP 8.5)") + 
  theme_map() +
  theme(strip.text = element_text(size = 7))

p1 | p2
```

## Key Differences I Found

**Looking at Summer (May-Aug):**

| Present | 2075 |
|---------|------|
| Some suitable habitat | Less suitable habitat |
| Moderate stress | More thermal stress |

**The pattern is similar because depth doesn't change, but temperature does!**

## What Does This Mean?

**For Lobsters:**

- Move **north** toward Maine and Canada
- Move to **deeper, cooler water**
- Summer becomes harder

**For Fishermen:**

- Fishing areas may shift north
- May need to adapt

## Summary

| Step | What I Did |
|------|------------|
| 1 | Got data from OBIS |
| 2 | Cleaned the data |
| 3 | Found what lobsters like |
| 4 | Trained MaxEnt model |
| 5 | Predicted present habitat |
| 6 | Predicted future (2075) |

## Main Finding

<div class="centered" style="font-size: 1.3em; margin-top: 80px;">
**Climate change will reduce summer habitat for American Lobsters.**
</div>

<div class="centered" style="font-size: 1.1em; margin-top: 30px;">
Lobsters will move **northward** and to **deeper water**.
</div>

## Thank You!

<div class="centered" style="font-size: 1.5em; margin-top: 100px;">
**Questions?**
</div>

<div class="centered" style="margin-top: 80px; font-size: 0.8em;">
*JP297Dj: Ocean Forecasting - AI, Ecology, and Data Justice*

*Colby College, January 27, 2026*
</div>
